Reviving the Lost Art of Combinator Graph Reduction

### 2023-01-24 experiments/superg/doc/paper/paper.tex
Related work.

\section{Related work}
supercombinators

GHC

Kiselyov's later work.

Edward Kmett's SIMD combinator stuff.

>>>>

\section{Related work}
\subsection{Super-combinators}
Super-combinators~\cite{hughes_super-combinators_1982} are the basis of many lazy functional runtimes~\cite{fairbairn_tim_1987,kieburtz_g-machine_1985,burn_spineless_1988,jones_implementing_nodate} developed after Miranda.
A supercombinator is a lambda expression of the form $\lambda x_1. \lambda x_2. ... \lambda x_n. E$, where $E$ has no free variables, and any lambda abstraction in $E$ is also a supercombinator~\cite{spj_impl}.
Any program in the lambda calculus may be transformed into supercombinators by \emph{lifting} free variables.
For example, we may consider $\lambda f.f\ (\lambda x.f\ x\ 2)$, which is not a super-combinator, because $f$ is free in the inner lambda.
To make this expression a super-combinator, we can pass $f$ explicitly:

\[
    \lambda f. f\ ((\lambda f.\lambda x. f\ x\ 2)\ f)
\]

This transformation is useful because the output of a supercombinator depends entirely on the value of its inputs, and thus its implementation may be described as a fixed set of (machine) instructions.

Programs translated into super-combinators have a code size that is linear in the size of the input in the worst case, an improvement over the worst-case quadratic size of bracket abstraction.

Perhaps more importantly, the reduction steps for a program based on super-combinators are larger.
A downside is that this makes the execution less lazy, but an important advantage is that the larger step size makes the overhead of the reduction loop smaller.
The small step size is an often noted issue with combinator reduction engines~\cite{spj_impl}, and a major reason why they were largely abandoned.

\subsection{Haskell}
Haskell is without a doubt the most popular lazy functional programming language today.
It too has its roots in the super-combinator approach.
The popular Glasgow Haskell Compiler is based on the Spineless Tagless G-Machine~\cite{jones_implementing_nodate}, which in turn is based on the Spineless G-Machine~\cite{burn_spineless_1988} and G-Machine~\cite{kieburtz_g-machine_1985}.
GHC compiles super-combinators into machine code ahead of time, which makes the resulting binaries self-contained and highly efficient.
In future work, we will compare the performance of our runtime against that of GHC.

\subsection{SIMD Combinator Reduction}
In his talk \emph{Combinator Revisited}~\cite{kmett_combinators_2018}, Kmett hypothesizes the use of SIMD instructions to perform combinator reductions.
The rationale is that combinators have a relatively simple implementation, and SIMD instructions are getting more and more powerful, so potentially they can be used to implement the reduction of multiple combinators in a single instructions.
Assuming the reduction of a single combinator uses 64-bit wide values, the same routine implemented with AVX-512 instructions could be used to perform as many as 8 reductions in parallel on a single core.
An eightfold improvement in performance is certainly promising, although it is unclear what level of parallelism a typical program may allow for.
A work-in-progress version of the compiler is available at \url{https://github.com/ekmett/coda}.
